---
biblio-style: jabes
bibliography: final.project.bib
documentclass: report
fontsize: 12pt
header-includes: \usepackage{mdwlist} \usepackage[compact]{titlesec} \usepackage{titling}
  \usepackage[font=small,labelfont=bf,tableposition=top]{caption} \usepackage{float}
  \floatstyle{plaintop} \restylefloat{table} \usepackage{lastpage} \usepackage{hyperref}
  \usepackage{colortbl} \usepackage{array} \hypersetup{backref,colorlinks=true} \usepackage{framed,color}
  \definecolor{shadecolor}{rgb}{0.95, 0.92, 0.88} \usepackage{graphicx} \usepackage{booktabs}
  \usepackage{fancyhdr} \usepackage[none]{hyphenat} \raggedright \usepackage{amsmath,
  amsthm, amssymb, bm} \usepackage{marginnote} \usepackage{subfig} \def\mygraphcaption{Here
  are my graphs.} \newlength{\mygraphwidth}\setlength{\mygraphwidth}{0.9\textwidth}
  \usepackage{listings}
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    highlight: tango
subparagraph: yes
---
  \lstset{
	basicstyle=\small\ttfamily,
	columns=flexible,
	breaklines=true}
	
  \pagestyle{fancy}
  \fancyhead[L]{\textbf{Don, Mel, Madhumithra}}
  \fancyhead[C]{}
  \fancyhead[R]{\textbf{STAT 823 Project}}
  \fancyfoot[L]{}
  \fancyfoot[C]{}
  \fancyfoot[R]{Page -\thepage- of \pageref{LastPage}}
  \fancypagestyle{plain}{\pagestyle{fancy}}
  \renewcommand{\headrulewidth}{2pt}
  \renewcommand{\footrulewidth}{2pt}
 
 \hypersetup{
	colorlinks   = true,
	citecolor    = blue,
	linkcolor    = black,
	urlcolor     = blue
  }
  
  \begin{titlepage}
   \begin{center}
       \vspace*{2cm}
        
       \vspace{0.5cm}
 
       \textbf{\textit{\LARGE Studying factors associated with prostate-specific antigen (PSA) levels in prostate cancer patients }}
 
       \vspace{0.5cm}
      
       \textbf{\Large STAT 823: Spring Class Project, 2021} 
       
        \vspace{0.5cm}
        
         \textbf{\large Don Ghysels, Melanie Valdez, Madhumithra Subramanian Karthikes}
        
       \vfill
 
       \vspace{0.7cm}
 
       \includegraphics[width=0.4\textwidth]{figures/ku}
 
       \large Department of Biostatistics and Data Science \\
       University of Kansas, USA \\
       `r format(Sys.time(), '%B %e, %Y')`
 
   \end{center}
\end{titlepage}

```{r setup, include=FALSE}
# load packages
library(knitr)
library(formatR)
library(stargazer)
library(xtable)
library(readxl)
library(ggplot2)
library(ggcorrplot)
library(gridExtra)
library(leaps)
library(car)
library(readr)
library(faraway)
library(epiDisplay)
knitr::opts_chunk$set(echo = FALSE)
options(digits = 5, width = 60, xtable.comment = FALSE)#,knitr.table.format="latex")
opts_chunk$set(tidy.opts = list(width.cutoff=60), tidy=TRUE)
out_type <- knitr::opts_knit$get("rmarkdown.pandoc.to")
```

 \setlength{\headheight}{45pt}
 
\thispagestyle{empty}
\newpage
\pagenumbering{roman}
\setcounter{page}{1}
\pagestyle{plain}
\tableofcontents
\cleardoublepage
\phantomsection
\listoftables
\phantomsection
\listoffigures
\newpage
\pagenumbering{arabic}

\section{Abstract}
Put your abstract here

\newpage

\section{Introduction}

Tracking the level of of the protein Prostate-specific antigen (PSA) in men diagnosed with prostate cancer is an an FDA approved method of tracking the disease (\citet{cit.nat.cancer.inst}). The objective of this study is to evaluate the association between \textit{ Serum Prostate-specific antigen level (mg/ml) \textbf{(PSA)}} and 7 clinical measurements in men with advanced prostate cancer.  The measurements inclue \textbf{Estimate of prostate cancer volume}, \textbf{Prostate weight}, \textbf{Age of patient}, \textbf{Amount of benign prostatic hyperplasia}, \textbf{Degree of capsular penetration}, \textbf{Seminal Vesical Invasion}, and the \textbf{Cleason Score}.

\section{Materials and Methods}

The dataset was obtained from the project materials provided for the stat823 course. A summary of the dependent and independent variables available is presented in Table \ref{table:model.variables}

\begin{table}[h]
\caption{Available independent and dependent variables}
\centering
\begin{tabular}{c c c c}
\hline
Variable & Units & Abbreviation & Type \\ [0.5ex] % inserts table
%heading
\hline
Serum prostate-specific antigen level & mg/ml & PSA & Continuous \textbf{(Dependent)} \\ % inserting body of the table
Estimate of prostate cancer volume & $cm^{3}$ & VOL & Continuous \\ 
Prostate weight  & gm & WT & Continuous \\
Age of patient & Years & AGE & Continuous \\
Amount of benign prostatic hyperplasia & $cm^{2}$ & HYP & Continuous \\
Degree of capsular penetration & cm & CAP & Continuous \\
Seminal Vesical Invasion & y/n & SEM & Categorical \\ 
Cleason Score & na & SCORE & Categorical \\ [1ex] % [1ex] adds vertical space
\hline %inserts single line
\end{tabular}
\label{table:model.variables}
\end{table}


\subsection{Statistical Analysis}

The data is analyzed using multiple linear regression.  

\subsubsection{Model Assumptions}

All inferences are conducted using $\alpha = 0.05$ unless stated otherwise.  No adjustments for multiplicity are made as this is an exploratory analysis. Discrete variables are summarized with proportions and frequencies.

\subsubsection{Primary Objective Analysis}

The primary objective analysis uses a Wilcoxon signed rank test to determine if there is a significant difference between cPRA scores at baseline and at week 16.  

\subsubsection{Goodness of Fit Test}

Describe here what goodness of fit statistics were used or will be used.

All of the statistical analyses in this document will be performed using `r R.version.string` [R](https://www.r-project.org/) \citep{r2018}.
R packages used will be maintained using the [packrat](http://rstudio.github.io/packrat/) dependency management system. 


```{r message=FALSE}
#load data and create the data frame
dataraw <- read_xlsx("../823data/pcancer.xlsx")
#create a new data frame and create factors for 
#categorical
data2 <- data.frame(dataraw)

data2$seminal <- factor(dataraw$seminal)
levels(data2$seminal) <- c("absence", "presence")
data2$score <- factor(dataraw$score)
levels(data2$score) <- c("6", "7", "8")

attach(data2)
```
\section{Results}

Describe your results here. Add subheadings if necessary.

\subsection{summary of statistics}

The summary statistics for the Prostate Cancer data are presented in Table \ref{tab:tab.summary}.  The standard deviation for the variable \textbf{WT} indicates outliers may be a concern.

```{r tab.summary, message=FALSE, results='asis'}
#create a table of summary statistics
summ.table <- summ(data2[2:9])$table
summ.table[1:8] <- c("PSA","VOL","WT","AGE","HYP","SEM","CAP","SCORE")
colnames(summ.table) <- c("Variable","Obs. Count","Mean","Median","SD","Min.","Max.")
xtable(summ.table,label="tab:tab.summary",caption = "Summary of statistics")
```

\section{Analysis of Covariates}

Correlation coefficients in Figure \ref{fig:fig.corr} indicate the best relationships are between \textit{PSA} and \textit{CAP} and between \textit{PSA} and \textit{VOL} 

The correlation between \textit{WT}, \textit{HYP} and \textit{AGE} are much weaker, indicating these variables may not be suitable as model covariates 

\subsection{Correlation Coefficients}

```{r, fig.corr, include=TRUE, message=FALSE, warning=FALSE, fig.cap="Correlation of available continuous covariates", fig.pos="H", fig.align='center', out.width="0.8\\linewidth"}
library(psych)
library(GGally)
#pairs.panels(data2[c(2:6,8)])

ggpairs(data2[c(2:6,8)])
```

\subsection{Data Distributions}

```{r include=TRUE}

dist.box.theme <-theme(plot.title = element_text(color="blue",
        size=10,
        face="bold",
        hjust = 0.5),
        axis.text.x = element_blank(),
        axis.ticks.x=element_blank(),
        panel.background = element_rect(fill = "white",
          colour = "blue",
          size = 1, 
          linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, 
          linetype = 'solid',
          colour = "lightgrey"))

box1 <- ggplot(data2, aes(y=cancerv,x=1)) + 
  geom_boxplot(outlier.color = "red") +
  geom_point(color="blue", size=1, position=position_jitter(0.05)) + 
  labs(x="",y="VOL")  + 
  dist.box.theme 
  
box2 <- ggplot(data2, aes(y=capsular,x=1)) + 
  geom_boxplot(outlier.color = "red") +
  geom_point(color="blue", size=1, position=position_jitter(0.05)) + 
  labs(x="",Y="CAP")  + dist.box.theme
  
box3 <- ggplot(data2, aes(y=hyperplasia,x=1)) + 
  geom_boxplot(outlier.color = "red") +
  geom_point(color="blue", size=1, position=position_jitter(0.05)) + 
  labs(x="",y="HYP") + dist.box.theme

box4 <- ggplot(data2, aes(y=weight,x=1)) + 
  geom_boxplot(outlier.color = "red") +
  geom_point(color="blue", size=1, position=position_jitter(0.05)) + 
  labs(x="",y="WT") + dist.box.theme

box5 <- ggplot(data2, aes(y=age,x=1)) + 
  geom_boxplot(outlier.color = "red") +
  geom_point(color="blue", size=1, position=position_jitter(0.05)) + 
  labs(x="",y="AGE") + dist.box.theme 

#---------- histogram-----------
fig.dist.hist.theme <-theme(plot.title = element_text(color="blue",
        size=10,
        face="bold",
        hjust = 0.5),
        axis.text.x = element_blank(),
        axis.ticks.x=element_blank(),
        panel.background = element_rect(fill = "white",
          colour = "blue",
          size = 1, 
          linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, 
          linetype = 'solid',
          colour = "lightgrey"))

hist1 <- ggplot(data2, aes(x=cancerv)) +
  xlab("VOL") + 
  #ggtitle("(a) VOL Histogram") +
  geom_histogram(binwidth = 2) + 
  fig.dist.hist.theme


hist2 <- ggplot(data2, aes(x=capsular)) +
  xlab("CAP") + 
  #ggtitle("(c) CAP Histogram") +
  geom_histogram(binwidth = 0.5)+ 
  fig.dist.hist.theme

hist3 <- ggplot(data2, aes(x=hyperplasia)) + 
  xlab("HYP") + 
  #ggtitle("(b) HYP Histogream") +
  geom_histogram(binwidth = 1)+ 
  fig.dist.hist.theme

hist4 <- ggplot(data2, aes(x=weight)) + 
  xlab("WT") +
  #ggtitle("(e) WT Histogram") +
  geom_histogram(binwidth = 0.8)+ 
  fig.dist.hist.theme

hist5 <- ggplot(data2, aes(x=age)) + 
  xlab("AGE") + 
  #ggtitle("(d) AGE Histogram") +
  geom_histogram(binwidth = 01)+ 
  fig.dist.hist.theme

```


Figure \ref{fig:fig.dist.VOL} through and Figure \ref{fig:fig.dist.AGE} summarize the variable distributions

```{r fig.dist.VOL, include=TRUE, fig.cap="Distribution of VOL",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
grid.arrange(ncol = 2, box1,hist1)
```

```{r fig.dist.CAP, include=TRUE, fig.cap="Distribution of CAP",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
grid.arrange(ncol = 2, box2,hist2)
```

```{r fig.dist.HYP, include=TRUE, fig.cap="Distribution of HYP",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
grid.arrange(ncol = 2, box3,hist3)
```

```{r fig.dist.WT, include=TRUE, fig.cap="Distribution of WT",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
grid.arrange(ncol = 2, box4,hist4)
```

```{r fig.dist.AGE, include=TRUE, fig.cap="Distribution of AGE",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
grid.arrange(ncol = 2, box5,hist5)
```


\begin{itemize}
  \item \textbf{VOL} The plots indicates the distribution may not have a normal distribution and shows two potential outliers should be checked.
  \item \textbf{CAP} The values of zero shown in the plot are assumed to be valid observations and not missing values.
  \item \textbf{HYP} The values of zero shown in the plot are assumed to be valid observations and not missing values.
  \item \textbf{WT}  The plot indicates two observations that should be examined as potential outliers. The plot shows two potential outliers
  \item \textbf{AGE} The plot indicates the distribution may be close to normal and does not show outliers
\end{itemize}


Figure \ref{fig:fig.dist.cat} shows the distribution of the of SCORE and SEM factors

```{r  fig.dist.cat, include=TRUE, fig.cap="distribution of factor variables",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
par(mfrow = c(1,2))
tab.sem <- tab1(data2$seminal, main = "Distribution of SEM")
tab.score <- tab1(data2$score, main="Distribution of SCORE")
```

```{r tab.sem.freq}
kable(tab.sem$output.table, caption = "SEM frequence plot")
```

```{r tab.score.freq}
kable(tab.score$output.table, caption = "SCORE frequence plot")
```


\subsection{Fitted Models}


```{r, tab.fitted.summary, message=FALSE, warning=FALSE, echo=FALSE}
lm.cancerv <- lm(data = data2, psa ~ cancerv)
lm.cancerv.log <- lm(data = data2, log(psa) ~ log(cancerv))
lm.hyperplasia <- lm(data = data2, psa ~ hyperplasia)
lm.capsular <- lm(data = data2, psa ~ capsular)
lm.weight <- lm(data = data2, psa ~ weight)
lm.weight.log <- lm(data = data2, log(psa) ~ log(weight))
lm.age <- lm(data = data2, psa ~ age)
lm.score6 <- lm(data = data2, psa ~ (score==6))
lm.score7 <- lm(data = data2, psa ~ (score==7))
lm.score8 <- lm(data = data2, psa ~ (score==8))
lm.seminal <- lm(data = data2, psa ~ seminal)

summ.cancerv <- summary(lm.cancerv)
summ.hyperplasia <- summary(lm.hyperplasia)
summ.capsular <- summary(lm.capsular)
summ.weight <- summary(lm.weight)
summ.age <- summary(lm.age)
summ.score8 <- summary(lm.score8)
summ.seminal <- summary(lm.seminal)

mse.cancerv <- mean(residuals(lm.cancerv)^2)
mse.hyperplasia <- mean(residuals(lm.hyperplasia)^2)
mse.capsular <- mean(residuals(lm.capsular)^2)
mse.weight <- mean(residuals(lm.weight)^2)
mse.age <- mean(residuals(lm.age)^2)
mse.score8 <- mean(residuals(lm.score8)^2)
mse.seminal <- mean(residuals(lm.seminal)^2)

summ.cancerv.v <- c(summ.cancerv$coefficients[c(2,4,6,8)],summ.cancerv$adj.r.squared,summ.cancerv$fstatistic[1])
summ.hyperplasia.v <- c(summ.hyperplasia$coefficients[c(2,4,6,8)],summ.hyperplasia$adj.r.squared,summ.hyperplasia$fstatistic[1])
summ.capsular.v <- c(summ.capsular$coefficients[c(2,4,6,8)],summ.capsular$adj.r.squared,summ.capsular$fstatistic[1])
summ.age.v <- c(summ.age$coefficients[c(2,4,6,8)],summ.age$adj.r.squared,summ.age$fstatistic[1])
summ.weight.v <- c(summ.weight$coefficients[c(2,4,6,8)],summ.weight$adj.r.squared,summ.weight$fstatistic[1])
summ.score8.v <- c(summ.score8$coefficients[c(2,4,6,8)],summ.score8$adj.r.squared,summ.score8$fstatistic[1])
summ.seminal.v <- c(summ.seminal$coefficients[c(2,4,6,8)],summ.seminal$adj.r.squared,summ.seminal$fstatistic[1])


summarydf <- data.frame("VOL"=summ.cancerv.v,
                        "HYP"=summ.hyperplasia.v,
                        "CAP"=summ.capsular.v,
                        "AGE"=summ.age.v,
                        "WT"=summ.weight.v,
                        "SCORE"=summ.score8.v,
                        "SEM"=summ.seminal.v)
rownames(summarydf) <- c("beta1","std Err","t-value","p-value","adj. R2","f-statistic")


for(i in 1:ncol(summarydf)){
  summarydf[[i]] <- sapply(summarydf[,i],function(x){if(x>0.0001){
                                                        return(format(x,digits=4))
                                                      }else{
                                                        return(format(x,digits=2))
                                                      }
                                                    })
}

knitr::kable(summarydf,label = "tab.fitted.summary", caption = "Goodness-of-fit")
                        
```


```{r fig.CAP.fitted, fig.cap="CAP Fitted Model", message=FALSE,fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}

par(mfrow=c(1,2),mar=c(5,4,4,3))
plot(y=data2$psa,x=data2$capsular, 
     xlab="PSA", 
     ylab="CAP")
box(col="blue")
mtext("CAP Fitted", side=3, line=0, cex=1, col="blue")
abline(lm.capsular,col="forestgreen")

plot(fitted(lm.capsular),residuals(lm.capsular), 
     xlab="residuals",
     ylab="fitted")
box(col="blue")
mtext("CAP Fitted vs Residuals", side=3, line=0, cex=1, col="blue")
abline(h = 0,col="forestgreen")

box("outer")
```

the fitted line in Figure \ref{fig:fig.CAP.fitted} indicates a poor linear relationship between PSA and Capsular.  The figure also shows indications the residuals will not support the non constant variance assumptions required for inference


```{r fig.VOL.fitted, fig.cap="VOL Fitted Model", message=FALSE,fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
par(mfrow=c(1,2),mar=c(5,4,4,3))
plot(y=data2$psa,x=data2$cancerv, 
     xlab="PSA", 
     ylab="VOL")
box(col="blue")
mtext("VOL Fitted", side=3, line=0, cex=1, col="blue")
abline(lm.cancerv,col="forestgreen")

plot(fitted(lm.cancerv),residuals(lm.cancerv), 
     xlab="residuals",
     ylab="fitted")
box(col="blue")
mtext("VOL Fitted vs Residuals", side=3, line=0, cex=1, col="blue")
abline(h = 0,col="forestgreen")

box("outer")

```

Figure \ref{fit:fig.VOL.dist} Shows very clear evidence of distribution problems.  The non constant variance test and shapiro-wilks test both indicate the data violates the assumption of constant variance and normal distribution}


```{r fig.WT.fitted, fig.cap="WT Fitted Model", message=FALSE,fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
par(mfrow=c(1,2),mar=c(5,4,4,3))
plot(y=data2$psa,x=data2$weight, 
     xlab="PSA", 
     ylab="WT")
box(col="blue")
mtext("WT Fitted", side=3, line=0, cex=1, col="blue")
abline(lm.weight,col="forestgreen")

plot(fitted(lm.weight),residuals(lm.weight), 
     xlab="residuals",
     ylab="fitted")
box(col="blue")
mtext("WT Fitted vs Residuals", side=3, line=0, cex=1, col="blue")
abline(h = 0,col="forestgreen")

box("outer")

```


```{r fig.AGE.fitted, fig.cap="AGE Fitted Model", message=FALSE,fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
par(mfrow=c(1,2))
plot(y=data2$psa,x=data2$age, 
     xlab="PSA", 
     ylab="AGE", 
     main="AGE Regression Fit")
abline(lm.age)

plot(fitted(lm.age),residuals(lm.age), 
     xlab="residuals",
     ylab="fitted",
     main = "AGE fitted vs residuals")
abline(h = 0)

```


```{r, HYP.fitted, echo=FALSE, fig.cap="HYP Fitted Model", message=FALSE,fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
par(mfrow=c(1,2))
plot(y=data2$psa,x=data2$age, 
     xlab="PSA", 
     ylab="HYP", 
     main="HYP Regression Fit")
abline(lm.age)

plot(fitted(lm.age),residuals(lm.age), 
     xlab="residuals",
     ylab="fitted",
     main = "HYP fitted vs residuals")
abline(h = 0)

```
\subsection{Transformations}

```{r echo=FALSE}
ncvp <-ncvTest(lm.cancerv)$p
swp <- shapiro.test(residuals(lm.cancerv))$p.value
```

The non-constant variance test and shapiro wilks test both indicate a problem with the distribution of the residuals.  The Shapiro-wilks p-value is `r swp` and the non-constant variance test p-value is `r ncvp` indicating the residuals are not normally distributed and the non-constant varaiance assumptions must be rejected. 

A log transformation of both psa and cancerv was found to generate a normal distribution of the residuals and also removed the problems with homoscedacity.

The regression plot and plot of the residuals vs. the fitted variables both indicate improvements in the distribution of the data

The p.values for shapiro wilks test and non-constant variance tests allow acceptance of the null hypothesis 

```{r log.VOL.fitted, echo=FALSE, include=TRUE, fig.cap="VOL Fitted AFter Log Transformation",fig.pos="H", fig.align='center', out.width="0.8\\linewidth"}

par(mfrow=c(2,2))
plot(y=log(data2$psa),x=log(data2$cancerv), 
     xlab="log(PSA)", 
     ylab="log(VOL)", 
     main="log(VOL) \n Regression Fit")

abline(lm.cancerv.log)

plot(fitted(lm.cancerv.log),residuals(lm.cancerv.log), 
     xlab="residuals",
     ylab="fitted",
     main = "Log(VOL)\n fitted vs residuals")
abline(h = 0)
```

```{r log.VOL.QQ, echo=FALSE, include=TRUE, fig.cap="VOL QQ Plot AFter Log Transformation",fig.pos="H", fig.align='center', out.width="0.8\\linewidth"}
qqnorm(residuals(lm.cancerv.log), main = "QQ Plot of log(VOL)")
qqline(residuals(lm.cancerv.log))
```

```{r tab4, echo=FALSE, include=TRUE, message=TRUE, results = 'asis'}
ncvp <-ncvTest(lm.cancerv.log)$p
swp <- shapiro.test(residuals(lm.cancerv.log))$p.value
cancerv.log.normdf <- data.frame(c(swp, ncvp))
row.names(cancerv.log.normdf) <- c("Shapiro-wilks p-value","ncv test p-value")
colnames(cancerv.log.normdf) <- c("values")
xtable(cancerv.log.normdf,label="tab:tab4",caption = "Shapiro-wilks and ncv test")
```

\section{Covariate Selection}

Examination of the p-values of the individual regression models in Table \ref{tab.fitted.summary} indicated \textbf{HYP}, \textbf{AGE}, and \textbf{WT} were not statistically significant.  The adjusted $R^{2}$ also indicated these variables did not have a strong linear relationship with \textbf{PSA},  \textbf{SCORE} was only statistically significant for \textbf{$SCORE_{=8}$}. \textbf{$SCORE_{=6}$} and \textbf{$SCORE_{=7}$} were not statistically significant. 

```{r tab.covariate.sel, results = 'asis'}

b <- regsubsets(log(psa) ~ log(cancerv)+seminal+capsular+score, data = data2)
regsub <- summary(b)
regsubdf <- data.frame(regsub$which)
#Assign columns names so we can look for each TRUE and buld a nice label
#Note: these column names are also used in the loop below.  
colnames(regsubdf) <- c("Intercept","VOL","SEM","CAP","SCORE7","SCORE8")

#dynamically build a label for the variables
#marked as TRUE to build the nice label
rowlabels <- vector(length = nrow(regsubdf))
for(i in 1:nrow(regsubdf)){
  r <- regsubdf[i,]
  lbl <- vector()
  if (regsubdf[i,"VOL"] == TRUE){
    lbl <-append(lbl, "VOL")
  }
  if (regsubdf[i,"CAP"] == TRUE){
    lbl <- append(lbl, "CAP")
  }
  if (regsubdf[i,"SEM"] == TRUE){
    lbl <- append(lbl, "SEM")
  }
  if (regsubdf[i,"SCORE7"] == TRUE){
    lbl <- append(lbl, "SCORE(7)")
  }
  if (regsubdf[i,"SCORE8"] == TRUE){
    lbl <- append(lbl, "SCORE(8)")
  }
  rowlabels[i] <- paste(lbl,collapse="+")
  
}

#build a new data frame using the row labels
regsubdf <- data.frame(rowlabels,"RSS" = regsub$rss, 
                       "AdjR2"=regsub$adjr2, 
                       "cp"=regsub$cp, 
                       "bic"=regsub$bic)
regsubdf[,2:5] <- round(regsubdf[,2:5],3)
xtable(regsubdf, digits = 2, label = "tab.covariate.sel")

```

Based on the lowest BIC, cp and the highest $R^{2}$, the following 3 covariates were selected for the final model: 

\begin{itemize}
  \item \textbf{VOL}
  \item \textbf{SCORE for value=8}
  \item \textbf{SEM}
\end{itemize}


test for linearity and variance for each variable
\section{final model}

\subsection{summary statistics}

\ref{tab:tab.lm.summary} indicates \textbf{VOL}, \textbf{$SCORE_{score=8}$} and \textbf{$SEM_{presence}$} are all statistically significant.

```{r tab.lm.summary,warning=FALSE, message=FALSE}
lm.multi <- lm(data=data2, log(psa) ~ log(cancerv) + score + seminal)
lm.multi.summary <- summary(lm.multi)
require(gtsummary)
library(gtsummary)

tbl_regression(lm.multi)
```

The response formula is as follows
$log(Y)$ = $\beta_{0} + \beta_{1}log(VOL) + \beta_{2}SCORE_{score=7} + + \beta_{3}SCORE_{score=8} +\beta_{4}SEM_{presence}$ 

where:  
$\beta_{0}$ = `r lm.multi.summary$coefficients["(Intercept)",1]`  
$\beta_{1}$ = `r lm.multi.summary$coefficients["log(cancerv)",1]`  
$\beta_{2}$ = `r lm.multi.summary$coefficients["score7",1]`  
$\beta_{3}$ = `r lm.multi.summary$coefficients["score8",1]`  
$\beta_{4}$ = `r lm.multi.summary$coefficients["seminalpresence",1]`  

\subsection{goodness-of-fit}


The adjusted $R^{2}$ value in table \ref{tab.summ.stat} indicates a moderate linear relationship

```{r tab.summ.stat, results = 'asis'}
mse.multi <- mean(residuals(lm.multi)^2)
radj.multi <- summary(lm.multi)$adj.r.squared

multi.summary.df <- data.frame("Values"=c(mse.multi,radj.multi))
row.names(multi.summary.df) <- c("MSE","Adjusted R2")
xtable(multi.summary.df,caption="Model Summary Statistics",label = "tab.summ.stat")
```

\subsection{influence}

```{r, tab.hat, results = 'asis'}
lm.multi.influence <- influence(lm.multi)
kable(head(sort(lm.multi.influence$hat,decreasing=TRUE),6), 
      col.names = c("Hat Influence"), caption = "Influence, top Hat values")
```

The hat values shown in \ref{tab.hat} shows observations 76 and 91 have the largest Hat values and the greatest influence

```{r fig.influence, include=TRUE, fig.cap="fitted vs residuals",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
multi.influence.tab <-influencePlot(lm.multi, id = list(n = 3))

```

```{r, tab.cooksd}
kable(multi.influence.tab[order(multi.influence.tab$CookD, decreasing = TRUE),c(1,3)],caption = "Top Cooks Distance")
```

```{r}
par(mfrow=c(2,2))
qqnorm(lm.multi.influence$coef[,2],main="VOL Influence")
qqnorm(lm.multi.influence$coef[,3],main="SCORE Influence")
qqnorm(lm.multi.influence$coef[,4],main="SEM Influence")
```



The cooks distance calculations in \ref{fig.cooksd} do not indicate these observations will cause a large change in the final model if observations 76 and 91 were removed.  

```{r}
par(mfrow=c(2,2))
plot(lm.multi.influence$coef[,2],main="VOL Influence")
plot(lm.multi.influence$coef[,4], main="SCORE Influence")
plot(lm.multi.influence$coef[,5], main="SEM Influence")
```
 
\subsection{residuals}

```{r fig.resid.vs.fitted, include=TRUE, fig.cap="fitted vs residuals",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}
#par(mfrow = c(1, 2))
plot(fitted(lm.multi),residuals(lm.multi), 
     main= "residuals vs fitted values"); 
abline(h = 0)
```

```{r fig.shapiro, include=TRUE, fig.cap="shapiro-wilks test of residuals",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}

shapiro.qqnorm(residuals(lm.multi), cex = 2)
```

```{r fig.ncv.test, include=TRUE, fig.cap="non-constant variance (Breusch-Pagan)test",fig.pos="H", fig.align='center', out.width="0.6\\linewidth"}

ncv <- (ncvTest(lm.multi))
ncv.out <- data.frame(c(ncv$ChiSquare,ncv$Df,ncv$p))
rownames(ncv.out) <- c("Chi-squre","Degrees of Freedom","p-value")
colnames(ncv.out) <- c("value")
kable(ncv.out,label = "fig.ncv.test")
```
The The p-value of `r ncv.out[3,]` from the non-constant variance test (shown table /ref{fig.ncv.test}) indicate the null hypothesis of constant variance should be accepted.   



```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```


