---
biblio-style: jabes
bibliography: lazbibreg.bib
documentclass: report
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{mdwlist} \usepackage[compact]{titlesec} \usepackage{titling}
  \usepackage[font=small,labelfont=bf,tableposition=top]{caption} \usepackage{float}
  \floatstyle{plaintop} \restylefloat{table} \usepackage{lastpage} \usepackage{hyperref}
  \usepackage{colortbl} \usepackage{array} \hypersetup{backref,colorlinks=true} \usepackage{framed,color}
  \definecolor{shadecolor}{rgb}{0.95, 0.92, 0.88} \usepackage{graphicx} \usepackage{booktabs}
  \usepackage{fancyhdr} \usepackage[none]{hyphenat} \raggedright \usepackage{amsmath,
  amsthm, amssymb, bm} \usepackage{marginnote} \usepackage{subfig} \def\mygraphcaption{Here
  are my graphs.} \newlength{\mygraphwidth}\setlength{\mygraphwidth}{0.9\textwidth}
  \usepackage{listings}
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    highlight: tango
  html_document:
    df_print: paged
subparagraph: yes
---
  \lstset{
	basicstyle=\small\ttfamily,
	columns=flexible,
	breaklines=true}
	
  \pagestyle{fancy}
  \fancyhead[L]{\textbf{abc, efg, hij}}
  \fancyhead[C]{}
  \fancyhead[R]{\textbf{STAT 823 Project}}
  \fancyfoot[L]{}
  \fancyfoot[C]{}
  \fancyfoot[R]{Page -\thepage- of \pageref{LastPage}}
  \fancypagestyle{plain}{\pagestyle{fancy}}
  \renewcommand{\headrulewidth}{2pt}
  \renewcommand{\footrulewidth}{2pt}
 
 \hypersetup{
	colorlinks   = true,
	citecolor    = blue,
	linkcolor    = black,
	urlcolor     = blue
  }
  
  \begin{titlepage}
   \begin{center}
       \vspace*{2cm}
        
       \vspace{0.5cm}
 
       \textbf{\textit{\LARGE Studying factors associated with prostate-specific antigen (PSA) levels in prostate cancer patients }}
 
       \vspace{0.5cm}
      
       \textbf{\Large STAT 823: Spring Class Project, 2021} 
       
        \vspace{0.5cm}
        
         \textbf{\large Don Ghysels, Melanie Valdez, Madhumithra Subramanian Karthikes}
        
       \vfill
 
       \vspace{0.7cm}
 
       \includegraphics[width=0.4\textwidth]{figures/ku}
 
       \large Department of Biostatistics and Data Science \\
       University of Kansas, USA \\
       `r format(Sys.time(), '%B %e, %Y')`
 
   \end{center}
\end{titlepage}

```{r setup, include=FALSE}
# load packages
library(knitr)
library(formatR)
library(stargazer)
library(xtable)
library(readxl)
library(ggplot2)
library(ggcorrplot)
knitr::opts_chunk$set(echo = TRUE)
options(digits = 5, width = 60, xtable.comment = FALSE)
opts_chunk$set(tidy.opts = list(width.cutoff=60), tidy=TRUE)
out_type <- knitr::opts_knit$get("rmarkdown.pandoc.to")
```

 \setlength{\headheight}{45pt}
 
\thispagestyle{empty}
\newpage
\pagenumbering{roman}
\setcounter{page}{1}
\pagestyle{plain}
\tableofcontents
\cleardoublepage
\phantomsection
\listoftables
\phantomsection
\listoffigures
\newpage
\pagenumbering{arabic}

\section{Abstract}
Put your abstract here

\newpage

\section{Introduction}
--------------  
Put all your introduction about the study topic here and cite some references. 

For example,  The textbook ``\textbf{Design and Analysis of Experiments}'' by \citet{voss} is required. 

Additional materials that you may wish to consult includ
\citet{casella}, 
\citet{kuehl}, 
\citet{shalabh}.

\subsectionsection{Data Sources}


\subsectionsection{Model Assumptions}


\subsectionsection{Data Sources}

\section{Analysis of Covariates}

```{r echo=FALSE}
dataraw <- read_xlsx("../823data/pcancer.xlsx")

data2 <- data.frame(dataraw)
data2$seminal <- factor(dataraw$seminal)
levels(data2$seminal) <- c("absence", "presence")
data2$score <- factor(dataraw$score)
levels(data2$score) <- c("6", "7", "8")
attach(data2)
```
\subsection{analysis of continuous Covariates}
```{r echo=FALSE}
summary(data2)
corr <- round(cor(data2[,c(3,4,5,6)]),4)
ggcorrplot(corr, hc.order = TRUE, type = "lower",
   lab = TRUE)
```

